Towards Practical First-Order Model Counting

### 2024-07-11 doc/paper/paper.tex
Remove unnecessary mentions of FCGs and (traditional) smoothing

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Recently, a new algorithm for FOMC
  was proposed that, instead of simply providing the final count, generates
  definitions of (possibly recursive) functions, which can be evaluated with
  different arguments to compute the model count for any domain size. However,
  the algorithm did not include base cases in the recursive definitions. This
  work makes three contributions. First, we demonstrate how to construct
  function definitions that include base cases by modifying the logical formulas
  used in the FOMC algorithm. Second, we extend the well-known circuit
  modification technique in knowledge compilation, known as smoothing, to work
  with the formulas corresponding to base cases and the recently proposed node
  types, such as domain recursion and constraint removal. Third, we introduce a
  compilation algorithm that transforms the function definitions into C++ code,
  equipped with arbitrary-precision arithmetic. These additions allow the new
  FOMC algorithm to scale to domain sizes over \num{9000} times larger than the
  current state of the art, as demonstrated through experimental results.
\end{abstract}

@@

% TODO: somehow indicate that source code will be available (a redacted link?)

@@

% TODO: must find a way to include the implementation in the submission

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for FCGs and the
formulas from \cref{step2} above required to preserve the correct model count.
Next, \cref{sec:cpp} elucidates how function definitions encoding a solution to
a WFOMC problem are compiled into C++ programs.\footnote{The possibility of
  compiling WFOMC solutions to C++ programs has been previously
  considered~\cite{DBLP:conf/kr/KazemiP16}. However, the extent of formulas that
  could be handled was limited.} Finally, \cref{sec:experiments} presents two
experiments comparing \Cranetwo{} with other state-of-the-art WFOMC algorithms.

@@

In the case of $n = 0$, there are three types of clauses to consider:
\begin{enumerate*}[label=(\roman*)]
  \item those that do not mention $\Delta$,\label[type]{type1}
  \item those in which every literal contains variables quantified over
  $\Delta$, and\label[type]{type2}
  \item those that have some literals with variables quantified over $\Delta$
  and some without.\label[type]{type3}
\end{enumerate*}
Clauses of \cref{type1} are transferred to the new formula $\phi'$ without any
changes. For clauses of \cref{type2}, $C'$ is empty, so these clauses are
filtered out. As for clauses of \cref{type3}, a new kind of smoothing is
performed, which will be explained in \cref{sec:smoothingbase}.

@@

\section{Smoothing}\label{sec:smoothing}

\emph{Smoothness} is a property originating in propositional knowledge
compilation, where it refers to the situation where all disjuncts in a
disjunction node contain the same atoms~\cite{darwiche2001tractable}.
\citeauthor{DBLP:conf/ijcai/BroeckTMDR11}~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}
extend this concept to first-order logic, introducing set-disjunction and
inclusion-exclusion nodes in addition to disjunction.

The purpose of smoothing is as follows. When compilation rules such as unit
propagation and inclusion-exclusion simplify a formula, certain ground atoms may
be eliminated (see \cref{example:basecasesmoothing} below). To properly account
for these atoms during counting, smoothing nodes (i.e., tautological clauses
such as $P(c) \lor \neg P(c)$) are added to the FCG in the appropriate location.

The remainder of this section presents an extension to the smoothing algorithm
of
\citeauthor{DBLP:conf/ijcai/BroeckTMDR11}~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
\Cref{sec:smoothingbase} explains the role of smoothing in the base-case-finding
algorithm described in \cref{sec:main}. \Cref{sec:smoothingfcg1} demonstrates
how to adapt smoothing to the compilation rules introduced by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}.

\subsection{Smoothing for Base Cases}\label{sec:smoothingbase}
% TODO: To be precise, line 7 of Algorithm 3 checks whether smoothing is
% necessary, and lines 8-9 execute it. If the condition on line 7 is not
% satisfied, the clause is not smoothed but rather completely omitted.

In this section, we motivate and describe \cref{line:smoothing1,line:smoothing2}
of \cref{alg:propagate}. Suppose that \Propagate is called with arguments
$(\phi, \Delta, 0)$, which means we are simplifying the formula $\phi$ by
assuming that the domain $\Delta$ is empty. Informally, if there is a predicate
$P$ in $\phi$ that has nothing to do with the domain $\Delta$, smoothing
preserves all occurrences of $P$ even if all clauses with $P$ become vacuously
satisfied. It is important to note that the approach presented in this section
is not unique. We explain it via the example below.
% TODO: In this context, "not unique" means that there are many different ways
% to adjust/rewrite the formula to achieve the goal of smoothing. This section
% (and Algorithm 3) presents one such approach.

@@

\subsection{Smoothing the FCG}\label{sec:smoothingfcg1}

Smoothing is a two-step process that involves propagating unit clauses upwards
(i.e., in the opposite direction of FCG arcs) and adding smoothing nodes to
account for missing atoms. In this section, we will
\begin{enumerate*}[label=(\roman*)]
  \item describe the relevant node types from previous work,
  \item explain how smoothing should work for these node types, and
  \item demonstrate the new smoothing techniques using two example FCGs.
\end{enumerate*}

Before discussing the proposed changes to smoothing, let us briefly review the
compilation rules and their corresponding node types introduced by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}.
\emph{Domain recursion} involves selecting a domain $\Delta$, introducing a new
constant $c \in \Delta$, and modifying the formula based on two possibilities
for each variable $x$ quantified over $\Delta$: $x = c$ or $x \ne c$. The
resulting node is denoted as $\DR(c \in \Delta)$. \emph{Constraint removal}
applies to formulas where each variable $x$ quantified over a domain $\Delta$ is
followed by the inequality constraint $x \ne c$. These formulas can be rewritten
so that each clause begins with $\forall x \in \Delta\text{.
}x \ne c \Rightarrow \ldots$ Constraint removal replaces $\Delta$ with a new
domain $\Delta'$ and removes the $x \ne c$ constraints. The resulting node is
denoted as $\CR(\Delta' \gets \Delta \setminus \{\,c\,\})$. Finally,
\emph{caching} detects when the input formula $\phi$ is equal to a previously
encountered formula $\psi$ except for having different domains. The resulting
node is denoted as $\Reff(\sigma)$, with $\sigma$ representing the substitution
mapping the domains of $\psi$ to their corresponding domains in $\phi$.

\paragraph{Stage~1 for Domain Recursion}
When visiting a $\DR(c \in \Delta)$ node, we replace each occurrence of
$\phi(c)$ or $\forall x \in \Delta\text{. } x \ne c \Rightarrow \phi(x)$ with
$\forall x \in \Delta\text{. }\phi(x)$ for each unit clause received from the
child node. This way, we ensure that if the subgraph below the domain recursion
node covers some of the ground atoms affected by domain recursion, it covers all
of them. If the relevant subgraph already covers those ground atoms, Stage~2
will not make any changes. Otherwise, smoothing nodes will be added below the
domain recursion node to account for the difference in the sets of ground atoms
assigned to the domain recursion node and its child node.

\paragraph{Stage~1 for Constraint Removal and Caching}
When visiting a $\CR(\Delta' \gets \Delta \setminus \{\,c\,\})$ node, we reverse
constraint removal by replacing each $\forall x \in \Delta'\text{. }\phi(x)$
with $\forall x \in \Delta\text{. }x \ne c \Rightarrow \phi(x)$. When visiting a
$\Reff(\sigma)$ node, we apply the substitution $\sigma$ to the unit clauses
from the child node.

\paragraph{Stage~2 for Domain Recursion}
We do not need to add smoothing nodes immediately below constraint removal or
caching nodes. However, for domain recursion, we follow a specific process.
\begin{enumerate}
  \item If the set of unit clauses assigned to the child node during Stage~1
        contains both $\phi(c)$ and $\forall x \in \Delta\text{.
        } x \ne c \Rightarrow \phi(x)$, we merge the two clauses into
        $\forall x \in \Delta\text{. }\phi(x)$. The only difference between
        these two clauses is that one has the constant $c$ while the other has a
        variable $x \ne c$.
  \item If necessary, we add smoothing nodes below the domain recursion node to
        account for the difference between the unit clauses assigned to the
        domain recursion node during Stage~1 and the unit clauses of the child
        node post-processed by the step above.
\end{enumerate}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw
    [$\land$,ellipse,draw,dashed,edge=dashed
    [$P(c)$,rectangle,draw,fill=green!20]
    [$\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \lor \neg P(x)$,rectangle,dashed,edge=dashed,draw,fill=blue!20]
    ]
    ]
  \end{forest}
  \caption{An artificial example of an FCG where a smoothing node needs to be
    added below a domain recursion node. The dashed nodes and arcs are added
    during Stage~2 of smoothing.}\label{fig:smoothing2}
\end{figure}

\begin{example}[An FCG that requires smoothing]
  \Cref{fig:smoothing2} depicts an FCG with a domain recursion node
  $\DR(c \in \Delta)$ followed directly by a single $P(c)$ node. In this
  scenario, Stage~1 assigns $\{\, \forall x \in \Delta\text{. }P(x) \,\}$ to the
  former node and $\{\, P(c) \,\}$ to the latter. As these two sets of unit
  clauses cover different ground atoms, Stage~2 adds a smoothing node to address
  $P(x)$ for all $x \in \Delta \setminus \{\, c \,\}$.
\end{example}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree={sn edges}
    [$\DR(c \in \Delta)$,name=dr,ellipse,draw,label={right:\textcolor{blue}{3. $\{\, \forall x \in \Delta\text{. }P(x) \,\}$}}
    [$\land$,ellipse,draw,label={left:\textcolor{blue}{2. $\{\, P(c) \,\}$}},label={right:\textcolor{blue}{6. $\{\, P(c), \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$P(c)$,rectangle,draw,fill=green!20,label={272:\textcolor{blue}{1. $\{\, P(c) \,\}$}}]
    [$\CR(\Delta' \gets \Delta \setminus \{\, c \,\})$,ellipse,draw,label={87:\textcolor{blue}{5. $\{\, \forall x \in \Delta\text{. }x \ne c \Rightarrow P(x) \,\}$}}
    [$\Reff(\Delta \mapsto \Delta')$,name=ref,ellipse,draw,label={below:\textcolor{blue}{4. $\{\, \forall x \in \Delta'\text{. }P(x) \,\}$}}]
    ]
    ]
    ]
  \end{forest}
  \caption{A smooth FCG based on \cref{example:overall}. For compactness, the
    arc between the caching node and the domain recursion node is not shown. The
    labels next to the nodes indicate the sets of unit clauses assigned to each
    node during Stage~1 and the order in which these assignments were made.
    Empty assignments that replace a set $S$ with $S$ itself are not
    included.}\label{fig:smoothing1}
\end{figure}

\begin{example}[A smooth FCG]\label{example:smooth}
  Stage~1 of the smoothing process is more complex in the case of the FCG shown
  in \cref{fig:smoothing1}. The unit clause $P(c)$ propagates to the conjunction
  node and is then generalised to $\forall x \in \Delta\text{. }P(x)$ by the
  domain recursion node. It continues to propagate to the caching node, changing
  its form to $(\forall x \in \Delta\text{.
  }P(x))[\Delta \mapsto \Delta'] \equiv \forall x \in \Delta'\text{. }P(x)$. The
  constraint removal node re-introduces the constraints, transforming the clause
  to $\forall x \in \Delta\text{. }x \ne c \Rightarrow P(x)$, which then
  propagates to the conjunction node, joining $P(c)$.

  During Stage~2, $P(c)$ and $\forall x \in \Delta\text{.
  }x \ne c \Rightarrow P(x)$ are combined into $\forall x \in \Delta\text{.
  }P(x)$. Since the resulting clause matches the clause assigned to the domain
  recursion node, the FCG is already smooth.
\end{example}

The algebraic interpretation of \cref{fig:smoothing1} is an equation $e$ that
defines a recursive function. The right-hand side of $e$ already encompasses
$P(c)$, and the smoothing algorithm correctly recognises that the recursive call
also encompasses $P(x)$ for all $x \in \Delta \setminus \{\, c \,\}$.

@@

In this work, we have presented several contributions. First, we have developed
algorithmic techniques to find the base cases of recursive functions generated
by the original \textsc{Crane} algorithm. Second, we have extended the smoothing
procedure of \textsc{ForcLift} and \textsc{Crane} to support base case formulas
and the compilation rules introduced by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. Third,
we have proposed an approach to compile function definitions into C++ programs
with support for arbitrary-precision arithmetic. Lastly, we have provided
experimental evidence demonstrating that \Cranetwo{} can scale to much larger
domain sizes than \textsc{FastWFOMC} while handling more formulas than
\textsc{ForcLift}.

@@

>>>>

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Recently, a new algorithm for FOMC
  was proposed that, instead of simply providing the final count, generates
  definitions of (possibly recursive) functions, which can be evaluated with
  different arguments to compute the model count for any domain size. However,
  the algorithm did not include base cases in the recursive definitions. This
  work makes three contributions. First, we demonstrate how to construct
  function definitions that include base cases by modifying the logical formulas
  used in the FOMC algorithm. Second, we extend the well-known circuit
  modification technique in knowledge compilation, known as smoothing, to work
  with the formulas corresponding to base cases. Third, we introduce a
  compilation algorithm that transforms the function definitions into C++ code,
  equipped with arbitrary-precision arithmetic. These additions allow the new
  FOMC algorithm to scale to domain sizes over \num{9000} times larger than the
  current state of the art, as demonstrated through experimental results.
\end{abstract}

@@

@@

@@

\Cref{fig:overview} outlines the workflow of the new algorithm. In
\cref{sec:main}, we describe how \CompileWithBaseCases finds the base cases for
recursive functions by:
\begin{enumerate*}[label=(\roman*)]
  \item identifying a sufficient set of base cases for each function,
  \item constructing formulas corresponding to these base cases,
  and\label[step]{step2}
  \item recursing on these new formulas.
\end{enumerate*}
Then, \cref{sec:smoothing} explains post-processing techniques for the formulas
from \cref{step2} above required to preserve the correct model count. Next,
\cref{sec:cpp} elucidates how function definitions encoding a solution to a
WFOMC problem are compiled into C++ programs.\footnote{The possibility of
  compiling WFOMC solutions to C++ programs has been previously
  considered~\cite{DBLP:conf/kr/KazemiP16}. However, the extent of formulas that
  could be handled was limited.} Finally, \cref{sec:experiments} presents two
experiments comparing \Cranetwo{} with other state-of-the-art WFOMC algorithms.

@@

In the case of $n = 0$, there are three types of clauses to consider:
\begin{enumerate*}[label=(\roman*)]
  \item those that do not mention $\Delta$,\label[type]{type1}
  \item those in which every literal contains variables quantified over
  $\Delta$, and\label[type]{type2}
  \item those that have some literals with variables quantified over $\Delta$
  and some without.\label[type]{type3}
\end{enumerate*}
Clauses of \cref{type1} are transferred to the new formula $\phi'$ without any
changes. For clauses of \cref{type2}, $C'$ is empty, so these clauses are
filtered out. As for clauses of \cref{type3}, a new kind of smoothing is
performed, which will be explained in \cref{sec:smoothing}.

@@

\section{Smoothing the Base Cases}\label{sec:smoothing}

\emph{Smoothing} is the process of modifying a circuit to re-introduce atoms
that might have been eliminated, thus ensuring that the circuit preserves the
correct (weighted) model
count~\cite{darwiche2001tractable,DBLP:conf/ijcai/BroeckTMDR11}. In this
section, we motivate and describe a similar process performed on
\cref{line:smoothing1,line:smoothing2} of \cref{alg:propagate}.

% TODO: To be precise, line 7 of Algorithm 3 checks whether smoothing is
% necessary, and lines 8-9 execute it. If the condition on line 7 is not
% satisfied, the clause is not smoothed but rather completely omitted.

Suppose that \Propagate is called with arguments $(\phi, \Delta, 0)$, which
means we are simplifying the formula $\phi$ by assuming that the domain $\Delta$
is empty. Informally, if there is a predicate $P$ in $\phi$ that has nothing to
do with the domain $\Delta$, smoothing preserves all occurrences of $P$ even if
all clauses with $P$ become vacuously satisfied. It is important to note that
the approach presented in this section is not unique. We explain it via the
example below.
% TODO: In this context, "not unique" means that there are many different ways
% to adjust/rewrite the formula to achieve the goal of smoothing. This section
% (and Algorithm 3) presents one such approach.

@@

@@

In this work, we have presented several contributions. First, we have developed
algorithmic techniques to find the base cases of recursive functions generated
by the original \textsc{Crane} algorithm. Second, we have extended the smoothing
procedure of \textsc{ForcLift} and \textsc{Crane} to support base case formulas.
Third, we have proposed an approach to compile function definitions into C++
programs with support for arbitrary-precision arithmetic. Lastly, we have
provided experimental evidence demonstrating that \Cranetwo{} can scale to much
larger domain sizes than \textsc{FastWFOMC} while handling more formulas than
\textsc{ForcLift}.

@@

% TODO:
% must clean pdf metadata before submitting
% make sure that it's not automatically deducible that I'm the author of Crane
% no type 3 fonts (including in the PDFs of figures)
% no overfull boxes (might be the figure)
% no embedded links (but URLs are fine as long as they're not clickable?)
% must not use 'input' command. instead, compile to a pdf
% figures should appear on the page where they are discussed or on the next page
% colors in figures must be cmyk, not rgb
% references must not have missing fields
